%!TEX root = ../latex/paper.tex

\section{Dynamic Type Checks in Grace}
\label{sec:method}

% evaluation cases

% high flexibility, no need for global constitency 


% design goals behind Grace / Moth
%As described in \cref{ssec:grace},
%Grace is an educational language and its performance profile
%is important for supporting effective teaching. 
%
%\mwh{It seems like a \textit{remarkably terrible} idea to be proposing
%  this as a contribution, especially when it's conflated and confused
%  with the performance contributions.}
%\kjx{we wont}
%
% motivation = give students low-overhead method to check program
The core of Grace's static type system is well described elsewhere\citep{JonesECOOP2016}; here we explain how these types can be understood
dynamically, from a student's or a programmer's point of view.
%\kjx{code for: no formalism, sorry.}
Following from the design goals behind Grace,
our motivation for this work
is to provide a flexible system 
to check consistency between an execution of a program
and its type annotations,
without significant impact on run-time performance.
A secondary goal is to have a design that can be implemented with
only a small set of changes to facilitate integration in existing systems.
%
% gradual typing drawback = slow
%\chg{Gradual typing systems are generally good candidates to achieve
%this goal.}{Gradual typing systems are good candidates,because they
%provide the gradual guarantee and the mechanisms for blame would be desirable.}
% Existing gradual typing systems 
% that address the goal of flexibility currently
% Unfortunately, they do not yet have the desired performance properties\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}.
% Specifically, they could encourage students to remove types to improve performance.
%


These goals are shared with much of the other work on gradual type
systems, but our context leads to some different choices. First, so
that students can see concrete examples of type errors, they should be
able to run their programs even if those programs are not type-correct%
---\ie Grace's static type checking is optional, and so an
implementation cannot depend on the correctness of a program's type
annotations.

Second, while checking Grace's type annotations
statically may be optional, checking them dynamically should
not be: any value that flows into a variable, argument, or result
annotated with a type must conform to that type annotation.
This means, the focus is not to devise a sound typing approach,
but one that ensure that the observed behavior matches
the one that one may expect considering the type annotations.

Third, adding type
annotations should not degrade a program's performance, or rather,
students should not be encouraged to improve performance by removing
type annotations.
However, reported errors should conform to the basic expectations one
may derive from the type annotations
and a strict interpretation of their shallow semantics.

Unfortunately, existing gradual type
implementations do not meet these goals, particularly regarding
performance; hence the ongoing debate about whether gradual typing is
alive, dead, or some state
inbetween\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}.
% Specifically, they could encourage students to remove types to improve performance.
% \mwh{This paragraph seems confused to me. They're good, but also they're bad. What is a ``gradual typing system'' here, and how does it relate to ``gradual typing'' and ``type systems'' as separate concepts?}
% \sm{the above change makes the desirable things about gradual typing more concrete. wrt ``gradual typing'' and ``type systems''. one is a general idea, and a concrete realization is a ``system''. I don't know whether this is how typing people talk. but this is how I see the distinction.
% we don't want to criticize the abstract idea, but concrete systems.}
% \kjx{I've rewritten this bit, particularly the last paragraph.}

%% TODO: (SM) need to look at this for revision:

% \kjx{I'm not sure if the goals should go here, or in the intro?}
% \kjx{now we have two different lists of goals here}
% \kjx{I'm more than half temped to just delete all of 3.0 --- but I've left
% it for now.}

% % KJX likes this bit, wrote it below, doesn't seem to belong there
% % 
% % Part of the philosophy of Grace is that the language should not force
% % students to annotate programs with types until they are ready, so that
% % teachers can choose whether to introduce types, early, late, or even
% % not a all.  Assuming 



% % summary of goals
% Instead, the design behind our type system is focused
% on maximizing performance and flexibility while
% preserving the ability for programmers to check the execution
% of their programs.

% % how we address flexible
% To address the goal of flexibility we propose a system that is optional,
% which enables the programmer to benefit from adding checks
% to be performed at run time,
% without the burden of needing to fully type each program.

% how we address min overhead (avoid blame)
% that enables the executed behavior of the program to be checked against
% the documented types without the
% (which leads to significant overhead in previous work).


\subsection{Design}


%% TODO: (SM) need to look at this for revision:
% \kjx{I worked though this mostly doing terminology,
% and perhaps moving from claims to descriptions.
% Does this make sense? is it any better?}

% JN killed
% To achieve the consistency checking with minimal run-time overhead,
% we propose a system with shallow structural type checks,
% which provide useful feedback without requiring a blame mechanism
% as in typical gradually typed approaches.

% Message sending
% As we described in \cref{ssec:grace}, Grace is message-sending-based language.

Our type checks for Grace are designed to be simple,
straightforward, and easy for students to understand:
\begin{itemize}
  \item types are shallow interfaces
  \item optional type annotations are checked at run time
  \item failing run-time type checks terminate execution
\end{itemize}

% Together these properties enable a flexible system
% with minimal overhead
% that developers can use to verify that programs
% execute with values that have the capabilities
% documented through their annotations
% while avoiding the burden of fully typing their program.

We illustrate how the type checks work in practice
in the context of an exercise where a student is developing
a program to record information about vehicles.
%
Grace is structurally typed\citep{graceOnward12}:
an object implements a type whenever it
implements all the methods required by a 
type rather than requiring types to be declared
explicitly.
In this context, methods match when they have the same name and arity.
A type expresses the requests
an object can respond to, for example whether a particular accessor is
available,  rather than a location in an explicit inheritance
hierarchy.

For example, our student can begin developing their vehicle
application by defining an object intended to represent a car
(\cref{lst:car-reg}, \cref{ex:object}) and write a method that, given
the car object, prints out its registration number (\cref{ex:method}).

\begin{lstlisting}[caption={The start of a simple program for tracking vehicle information.},float,label=lst:car-reg,escapechar=|,columns=flexible]
def car = object {|\label{ex:object}|
    var registration is public := "JO3553"
}

method printRegistration(v) {|\label{ex:method}|
    print "Registration: {v.registration}"
}
\end{lstlisting}

Next, the student could decide to ensure that any object passed to the
\code{printRegistration} method will respond to the
\code{registration} request.  To get this support, the student first
defines the structural type \code{Vehicle}\citep{theCleanVehicle}
naming just that method (\cref{ex:vehicle},
\cref{ex:adding-type:vehicle}), and then annotates the
\code{printRegistration} method's argument with that type
(\cref{ex:vehicle}, \cref{ex:adding-type}). This ensures the student
will be alerted as soon as a value that does not conform to this
expected type is passed into the \code{printRegistration} method,
rather than crashing somewhere in the middle of the
implementation of the \code{print} method.


\begin{lstlisting}[label={ex:vehicle},caption={Adding a type annotation to a method parameter.},escapechar=|,columns=flexible]
type Vehicle = interface { |\label{ex:adding-type:vehicle}|
    registration    
}

method printRegistration(v: Vehicle) { |\label{ex:adding-type}|
    print "Registration: {v.registration}"
}
\end{lstlisting}

% \sm{you already said the following}
% By adding the type annotation to the program,
% the student can now be sure the \code{getReg} method
% will only be invoked when its arguments can respond to this message.

%\paragraph{Flexibility.}
%\paragraph{Semnatics}

While Grace's static type system supports full static type
checking\citep{graceOnward12}, Grace's specification requires dynamic
type tests to be \emph{shallow}, that is, they check only for the
presence and arity of methods in an object, rather than also checking
conformance of argument and result types.  This is to ensure that the
presence or absence of type annotations does not affect the execution
of a program, for the reason originally outlined by
\citet{Boyland2014}, thus maintaining a version of the gradual
guarantee.  The resulting semantics are more-or-less equivalent to
type-tag soundness\citep{Greenman2018}---the difference being that
where type-tag soundness supports shallow \emph{nominal} type checks,
we support shallow \emph{structural} type checks.

% The notion of our type system being shallow means 
% that members of a type are untyped.
% In particular, our design does not have information on parameter types
% and return types for types' members.
% This design implies that blame tracking is not needed,\mwh{Um.}
% because the types are less detailed than in other systems.\mwh{I would hope that a competent reviewer would't fall for this\ldots}\sm{ok, what can we do about this? what's the precise issue here?
% The idea is that we do not need blame, because we do not have the type casts
% from blame-supported gradual systems.
% The difference is, I think that there is nothing we can not check immediately,
% and error as soon as there is a type.
% I believe the examples are usually higher-order functions/blocks.
% We only check the arity. So, there is no type cast wrapping necessary,
% when we pass it into a method that would expect another type of block arguments.
% (because passing in should not error, the error should only happen when
% the block is used in these systems, right?)}
% One reason for this design is that it avoids the overhead incurred
% by tracking type assumptions for precise blame attribution.



% Another aspect of this design is
% that our shallow approach to types
% allows for more flexible use 
% without requiring type parameters, \ie, generics.
For example, in \cref{ex:complex}, the student
creates two cars objects
(\cref{ex:personal-car,ex:government-car}),
that conform to
an expanded \code{Vehicle} type (\cref{ex:new-vehicle}).
Note that each version of the \code{registerTo} method
declares a different type for its parameter
(\cref{ex:personal-car:registerTo,ex:government-car:registerTo}).
% Depending on its semantics,
% a less-shallow type checking system could throw an error
% due to the inconsistency.
% In contrast
% our approach allows the student to execute the program 
% despite its inconsistency, 
% while preserving the guarantee that any value found to be
% inconsistent with the annotation will result in
% a termination by a typing error.
When the student runs this program, both \code{personal\-Car} and
\code{governmentCar} can be assigned to \code{Vehicle} because that
check considers only that the vehicle has a \code{registerTo} method,
but not the required argument type of that method.
At \cref{ex:invoke-register-to} the student can attempt to register a
government car to a person: only when the method is invoked
(\cref{ex:government-car:registerTo}) the dynamic type test on the
argument will fail (the object that is passed in is not a
\code{Department}) even though the body of the \code{registerTo} method
does not rely on the \code{code} method that the \code{Department}
annotation requires of the argument.
The intention here is to ensure that the run-time values
match the given types in a strict and eager sense,
which is the intuition we derive from types having a constraining meaning
on values.

% \sm{don't say anything about stuff we don't do. this is not really necessary,
% and I don't think this is necessarily correct, depending on specific type systems}
% A full structural type-checking system would alert the student to
% the inconsistency described above statically,
% perhaps displaying an error in their idea.
% While the alert is correct in that their is an inconsistency,
% such a system is less flexible.

% \sm{I would just say: possible inconsistency between elements can be approached
% step wise, when necessary as the sophistication and completeness of the program improves}
% Our system offers a higher level of flexibility in that the student
% may still execute the program,
% which may indeed be only partially developed and remain globally inconsistent.
% The student may be satisfied that a particular test passes successfully, and
% is then free to address the inconsistency in later development.


\begin{lstlisting}[caption={A program in development with inconsistent
    types.},escapechar=|,label={ex:complex},float,floatplacement=htb,columns=flexible]
type Vehicle = interface { |\label{ex:new-vehicle}|
    registration
    registerTo(_)
}

type Person = interface { name }
type Department = interface { code }

var personalCar : Vehicle := |\label{ex:personal-car}|
  object {
    var registration is public := "DLS018"
    method registerTo(p: Person) {|\label{ex:personal-car:registerTo}|
      print "{p.name} registers {self}"
    } 
  }

var governmentCar : Vehicle := |\label{ex:government-car}|
  object {
    var registration is public := "FKD218"
    method registerTo(d: Department) { |\label{ex:government-car:registerTo}|
      print "some department {self}"
    }
  }

governmentCar.registerTo( |\label{ex:invoke-register-to}|
  object {
    var name is public := "Richard"
  }
)
\end{lstlisting}


%\paragraph{Termination by Type Error.}
% \label{sec:term-type-error}

% When executing a program without types, there are three possible outcomes.
% Either the program (1) terminates successfully,
% (2) terminates with an exception, or 
% (3) the execution diverges, \ie, it does not terminate.
% Using our approach, a
% fourth outcome is possible: termination with a type error.
% Our implementation checks  every type annotation
% on the values of arguments before invoking a method, 
% on the value returned by a method when it returns, and
% before any assignment to and read from a variable 
% (either local to a method or belonging to an object). 

% The checks are performed eagerly%
% ---as soon as they are encountered during execution---%
% and cause the execution to terminate with a typing error 
% when a value fails to implement its expected type.

% % kjx deleted this summary because it doesn't add anything.
% % 
% % % Summarize the design section
% % \paragraph{Summary}
% % Our type checking approach enables developers to express
% % the capabilities of objects throughout different components of 
% % their programs.
% % Our representation of types is shallow,
% % in that a type expresses only the set of members an object
% % should offer, while excluding any further typing information. 
% % The shallow design enables a flexible use of structural types without
% % requiring type parameters.
% % We abort program execution as soon as
% % a type annotation is inconsistent with a concrete value. 
% % With these design choices, our system offers a
% % mechanism to check for well-typed executions%
% % ---rather than well-typed programs---%
% % without negatively affecting the performance profile.
% % % This both lowers overhead for the developer and,
% % % helps us to avoid the overhead of tracking blame
% % % as seen among previous works.

\subsection{Implementation} 
\label{ssec:implementation} 

%This section gives an overview of a possible implementation
%based on an abstract-syntax-tree (AST) interpreter.

We have implemented shallow dynamic structural type checks 
by extending the Moth abstract-syntax-tree (AST) interpreter for
Grace (\cref{ssec:moth}).
%
% We developed our implementation as an extension to Moth.
% As described earlier in \cref{ssec:moth},
% Moth is an AST-based interpreter on top of the Graal VM.
%
% It is optimizes itself based on for instance run-time type information.
%
%
%Based on \cref{sec:term-type-error},
Our approach needs to check types of values at run-time:

\begin{itemize}
\item the values of arguments are checked after a method is requested, 
      but before the body of the message is executed,
\item the value returned by a method is checked after its body is executed, and
\item the values of variables are checked
      whenever written or read by user code.
\end{itemize}

% \kjxdone{why both reading and writing?  answer - doesn't matter?}

One of the goals for our approach to dynamic typing was to keep
the necessary changes to an existing implementation small,
while enabling optimization in highly efficient language runtimes.
%
In an AST interpreter, we can implement this approach by attaching the
checks to the relevant AST nodes: the expected types for the argument
and return values can be included with the node for requesting a
method, and the expected type for a variable can be attached to the
nodes for reading from and writing to that variable.  In practice, we
encapsulate the logic of the check within a new type-checking AST
node.  Moth's front end was adapted to parse and record type
annotations and attach instances of this checking node as children of the
existing method, variable read, and variable write nodes.


% The check node is detailed in \cref{ssec:optimization} to discuss
% relevant optimizations.

%

The check node uses the internal representation of a Grace type
(cf. \cref{ex:type}, \cref{ex:type:check}) to test whether an observed
object conforms to that type. 
% These \code{Type} objects are created by
% Grace \code{interface} expressions, and also help
% to support Grace's pattern matching facilities \cite{gracePatternsDLS12}.
An object satisfies a type if all members required by the type are provided
by the object (\cref{ex:type:satisfied}).


\begin{lstlisting}[label={ex:type},escapechar=|,caption={Sketch of a \code{Type} in our system and its \code{check()} semantics.},float,floatplacement=htb,columns=flexible]
class Type:
  def init(members):
    self._members = members

  def is_satisfied_by(other: Type): |\label{ex:type:satisfied}|
    for m in self._members:
      if m not in other._members:
        return False
    return True

  def check(obj: Object):
    t = get_type(obj)
    return self.is_satisfied_by(t) |\label{ex:type:check}|
\end{lstlisting}


\subsection{Optimization}
\label{ssec:optimization}

\smtodo{would be nice to measure all of these optimizations...}

There are two aspects to our implementation that are critical for a minimal overhead solution:

\begin{itemize}
  \item specialized executions of the type checking node, along with guards to protect these specialized version, and
  \item a matrix to cache sub-typing relationships to eliminate redundant executions.
\end{itemize}
 
%Here we discuss each of the aspects in more detail.

\begin{lstlisting}[label={ex:typenode},escapechar=|,caption={An illustration of the type checking node that support type checking},float,floatplacement=htbp,columns=flexible]
class TypeCheckNode(Node):

  expected: Type
  record: Matrix

  @Spec(static_guard=expected.check(obj))
  def check(obj: Number): |\label{ex:typenode:number}|
    pass

  @Spec(static_guard=expected.check(obj))
  def check(obj: String): |\label{ex:typenode:string}|
    pass

  ...

  @Spec(
      guard=obj.shape==cached_shape,
      static_guard=expected.check(obj))
  def check(obj: Object, @Cached(obj.shape) cached_shape: Shape): |\label{ex:typenode:object}|
    pass
  
  @Fallback
  def check_generic(obj: Any): |\label{ex:typenode:generic}|
    T = get_type(obj)
    
    if record[T, expected] is unknown: |\label{ex:typenode:matrix}|
      record[T, expected] =
          T.is_subtype_of(expected) |\label{ex:typenode:reuse}|

    if not record[T, expected]:
      raise TypeError(
          "{obj} doesn't implement {expected}")
\end{lstlisting}

The first performance-critical aspect to our implementation
is the optimization of the type checking node.
We rely on Truffle and its TruffleDSL\citep{humer2014domainspecific}.
This means we provide a number of special cases,
which are selected during execution based on the observed concrete 
kinds of objects.
A sketch of our type checking node using a pseudo-code version of the DSL
is given in \cref{ex:typenode}.
A simple optimization is for well known types such as
numbers (\cref{ex:typenode:number}) or strings (\cref{ex:typenode:string}).
The methods annotated with \code{@Spec} (shorthand for \code{@Specialization})
correspond to possible states in a state machine that is generated by the
TruffleDSL.
Thus, if a check node observes a number or a string,
it will check on the first execution only that the expected type,
\ie, the one defined by some type annotation,
is satisfied by the object by using a \code{static\_guard}.
If this is the case, the DSL will activate this state.
For just-in-time compilation, only the activated states and their normal guards are considered.
A \code{static\_guard} is not included in the optimized code.
If a check fails, or no specialization matches, the fallback,
\ie, \code{check\_generic} is selected (\cref{ex:typenode:generic}),
which may raise a type error.

For generic objects, we rely on the specialization on \cref{ex:typenode:object},
which checks that the object satisfies the expected type.
If that is the case, it reads the shape of the object (cf. \cref{ssec:moth}) at specialization time,
and caches it for later comparisons.
Thus, during normal execution,
we only need to read the shape of the object and then compare it to the cached one
with a simple reference comparison.
If the shapes are the same, we can assume the type check passed successfully.
Note that shapes are not equivalent to types.
However, shapes imply the set of members of an object, and thus,
do imply whether an object fulfills a type.

The other performance-critical aspect to our implementation
is the use of a matrix to cache sub-typing relationships.
The matrix compares types against types,
featuring all known types along the columns and the same types again along the rows.
A cell in the table corresponds to a sub-typing relationship:
does the type corresponding to the row implement
the type corresponding to the column?
All cells in the matrix begin as unknown and, as
encountered in checks during execution, we populate the table.
If a particular relationship has been computed before
we can skip the check and instead recall the previously-computed value
(\cref{ex:typenode:reuse}).
Using this table we are able to eliminate the redundancy of evaluating
the same type to type relationships across different checks in the program. To reduce redundancy further we also unify types in a similar way to Java's string interning; 
during the construction of a type we first check to see if the same
set of members is expressed by a previously-created type and, if so,
we avoid creating the new instance and provide the existing one instead.

Together the self-specializing type check node and the matrix-based record
ensure that our implementation eliminates redundancy, and
consequently, we are able to minimize the run-time overhead of our system. 

% The Graal compiler 
% Moth needs to query the shape of arguments to evaluate them.
% Moth then needs to query the shape again to run the specializations' guards.
% Once profiled Graal can remove this redundancy by eliminating
% reusing the first query.
% During this compilation Graal also employs other optimizations 
% -- sub-expression elimination, inlining, propagation and folding --\tabularnewline
% and, as we demonstrate next in \cref{sec:evaluation},
% produces a compiled version of the method with minimal overhead incurred from
% the type check.
